---
title: 'DIY Metrics:  Net Ratings (ish)'
author: Matthew Avery
date: '2019-03-09'
slug: diy-metrics-net-ratings-ish
categories:
  - basketball
  - statistics
  - metrics
  - sports
  - R
tags: []
description: ''
thumbnail: ''
---
```{r, include = F}
suppressMessages(library(lubridate))
suppressMessages(library(tidyverse))

tmp <- read_rds("data/diy-metrics/team-events.rds")
```

<a href = "/post/diy-metrics-five-man-unit-plus-minus/">Last time on DIY Metrics</a>, we calculated five-man-unit plus/minus ratings from scratch. If we want to use measures like this to compare performance for these groups of players, its important to consider how much game time we have for each unit. There's a relevant discussion to be had about whether "number of posessions" or "elapsed time" is the best way to compare these groups, (IMO, it depends on what specific question you're trying to answer with your metric) but today we'll avoid that discussion and normalize over time because it's easier.

### Data Cleaning

Our first step is to build out our workflow to allow us to do more interesting things. Thus far, we've been starting from scratch each time. This isn't typically how you'd work on a project. The workflow we'll show today is better for longer term projects. We'll start by building up a script that generates a "cleaned" version of our raw data set (from <a href = "https://www.nbastuffer.com/analytics101/playbyplay-data/">NBAstuffer</a>).

Here's our script for data cleaning:

```{r, eval = F}

library(tidyverse)

tbraw <- read_csv("data/sample-combined-pbp-stats.csv") 

teamsandplayers <- tbraw %>% 
  filter(elapsed > 0) %>% #eliminates weird cases like opening jump ball
  select(player, team) %>% 
  filter(!is.na(player)) %>% 
  distinct() 

tb <- tbraw %>% 
  mutate(hometeam = get_team(h1, teamsandplayers),
         awayteam = get_team(a1, teamsandplayers),
         pointchange = map_lgl(points, score_changed)) 


allteams <- tb %>% 
  filter(elapsed > 0) %>% #eliminates weird cases like opening jump ball
  select(team) %>% 
  filter(!is.na(team)) %>% 
  distinct() 

tmp <- left_join(allteams, get_team_events(allteams, tb))
```

Simple, right? Plus a lot of those function calls look familiar from previous installments. Well, except for one. The main work-horse here is `get_team_events`, which is worth looking at in more detail. (I've included all the helper functions I used at end of the post if you're interested in them or want to try to run this code on your own.)

```{r, eval = F}
get_team_events <- function(whichteam, tb){
  
  out <- NULL
  for(i in seq_along(whichteam$team)){
    
    thatteam <- whichteam$team[i]
    
    teamtbl <- tb %>% 
      mutate(currentteam = thatteam) %>% 
      filter(hometeam == currentteam | awayteam == currentteam) %>% 
      fiveplayers()
    
    thisteamsplayers <- get_this_teams_players(teamtbl, thatteam)
    
    out <- teamtbl %>% 
      add_subsections(thisteamsplayers) %>% 
      mutate(netpoints = pmap_dbl(list(points, currentteam, team), get_net_points)) %>% 
      select(-team) %>% rename(team = currentteam) %>% 
      bind_rows(out)
    
  }
  out %>% group_by(team) %>% nest(.key = `team events`)
  
}
```

For each team in the data set, this function creates a specific table that contains all events that team participated in, and tranforms the outcomes (e.g., whether a basket is a "positive" or "negative" to the perspective of that team. If it does look kinda familiar, it's an updated version of `get_team_scores` from last week. The key difference is the `add_subsections` subfunction, which divides the game into subsections with identical personnel (from the team of interest). This can be useful if we want to aggregate over these time periods.

The last step in this script is to save the cleaned data set for future retrieval and use:

```{r, eval = F}
write_rds(tmp, "clean-data/team-events.rds")
```

I save it as an rds file here to preserve the nesting behavior of the `tbl_df` we've just built. You can also just saves it as a .csv if you want the more general formatting. 

### Getting the plus/minus scores and times

With our nicely cleaned data set, it's now pretty simple to get the information we want. 

First, we read in our data:

```{r, eval = F}
library(lubridate)
library(tidyverse)

tmp <- read_rds("clean-data/team-events.rds")
```

Next, we'll need a new function:


```{r}
get_fiveman_plusminus <- function(tb, team){
  
  tb %>% 
    mutate(team = team) %>% 
    mutate(pl = hms(play_length)) %>% 
    group_by(p1, p2, p3, p4, p5) %>% 
    summarise(`Plus/Minus` = sum(netpoints, na.rm = T),
              unittime = sum(pl)) 
  
}

```

This function is designed to work with `purrr::map2` on the nested data set we just built. To get the time values, we use the `play_length`  column from our data set along with `lubridate::hms`, which takes the column and tells `R` how to process it properly. I found this worked so well that I want to pause to illustrate it quickly:

```{r}
tmp %>% 
  filter(team == "BOS") %>% 
  select(`team events`) %>% 
  unnest %>% 
  select(play_length) %>% 
  mutate(pls = hms(play_length)) #  <-- critical line of code from lubridate
```

Boy, howdy that's easy! And the new column will do intuitive things when, for example, you use functions like `sum` over it. 


Now let's build our plus-minus data and check it out!
```{r}

pm <- tmp %>%
  mutate(`Five-man Plus/Minus` = map2(`team events`, team, get_fiveman_plusminus))

pm %>% filter(team == "BOS") %>% select(`Five-man Plus/Minus`) %>% unnest()
```

This is basically where we were last time, except now we've got the total duration of game time each of these units played. This also has the values broken out by "game section" as we defined it above, so we'll have to include a `summarise` step to aggregate these. Also note that "0 second" periods. This is due to a quirk with the data set, which codes each substitution individually. When two players check in simultaneously, there's a "0 second duration" section when only one player has checked in. So basically we want to ignore those sections, and we do so with a `filter(unittime > 0)` step.

By taking those times, summing over we can normalize unit performance to a per-36-minute basis:
```{r}

pm %>% filter(team == "BOS") %>% select(`Five-man Plus/Minus`) %>% unnest() %>% 
  group_by(p1, p2, p3, p4, p5) %>% 
  filter(unittime > 0) %>% 
  summarise(tpm = sum(`Plus/Minus`),
            ttime = sum(unittime)) %>% 
  mutate(normpm = tpm * 3600 / ttime)

```

So there you have it! These numbers look wonky becayse we're still just looking at one or two game's worth of data for each team.  Suppose I'll have to break down and pay for the full season's worth of data at some point so this looks more reasonable....

But good enough for now!

#### Helper functions (data cleaning)

These are the helper functions used in the `get_team_events` function which does the main data cleaning from above:

```{r}

get_team <- function(playernames, teamsandplayers){
  
  teamsandplayers$team[match(playernames, teamsandplayers$player)]
  
}

score_changed <- function(points){
  
  !is.na(points) && points > 0
}

get_net_points <- function(points, histeam, team){
  if(is.na(team))
    return(NA) else
      if(histeam == team)
        return(points) else
          return(points * -1)
}


add_subsections <- function(teamtbl, thisteamsplayers){
  
  subsec <- rep(0, nrow(teamtbl))
  
  counter <- 1
  for(i in 1:nrow(teamtbl)){
    if(!is.na(teamtbl$entered[i]) && teamtbl$entered[i] %in% thisteamsplayers$player) counter <- counter + 1
    subsec[i] <- counter
  }
  
  teamtbl$subsec <- subsec
  teamtbl
  
}


get_this_teams_players <- function(teamtbl, thatteam){
  
  teamtbl %>% 
    filter(elapsed > 0,             #eliminates weird cases like opening jump ball
           team == thatteam) %>% 
    select(player, team) %>% 
    filter(!is.na(player)) %>% 
    distinct()
}

fiveplayers <- function(tb){
  
  tb1 <- tb %>% filter(currentteam == hometeam) %>% 
    rename(p1 = h1, p2 = h2, p3 = h3, p4 = h4, p5 = h5, o1 = a1, o2 = a2, o3 = a3, o4 = a4, o5 = a5)
  
  tb2 <- tb %>% filter(currentteam == awayteam) %>% 
    rename(p1 = a1, p2 = a2, p3 = a3, p4 = a4, p5 = a5, o1 = h1, o2 = h2, o3 = h3, o4 = h4, o5 = h5)
  
  bind_rows(tb1, tb2)
  
}

```

